% filename: pow_x_n.tex

\problemsection{Pow(x, n)}
\label{chap:Pow_x_n}
\marginnote{\href{https://leetcode.com/problems/powx-n/}{[LeetCode Link]}\index{LeetCode}}
\marginnote{\href{https://www.geeksforgeeks.org/write-a-c-program-to-calculate-powx-n/}{[GeeksForGeeks Link]}\index{GeeksForGeeks}}
\marginnote{\href{https://www.interviewbit.com/problems/power-of-two-integer/}{[InterviewBit Link]}\index{InterviewBit}}
\marginnote{\href{https://app.codesignal.com/challenges/powx-n}{[CodeSignal Link]}\index{CodeSignal}}
\marginnote{\href{https://www.codewars.com/kata/powx-n/train/python}{[Codewars Link]}\index{Codewars}}

The \textbf{Pow(x, n)} problem is a fundamental algorithmic challenge that involves computing the value of \(x\) raised to the power of \(n\) (\(x^n\)). This problem tests one's understanding of mathematical principles, algorithm optimization, and efficient implementation techniques. Mastery of this problem is essential for tackling more complex computational tasks in areas such as numerical analysis, computer graphics, and scientific computing.

\section*{Problem Statement}

Implement the function \texttt{pow(x, n)} which calculates \(x\) raised to the power of \(n\) (\(x^n\)), where \(x\) is a floating-point number and \(n\) is an integer.

\textbf{Function signature in Python:}
\begin{lstlisting}[language=Python]
def myPow(x: float, n: int) -> float:
\end{lstlisting}

\section*{Examples}

\textbf{Example 1:}

\begin{verbatim}
Input: x = 2.00000, n = 10
Output: 1024.00000
Explanation: 2^10 = 1024
\end{verbatim}

\textbf{Example 2:}

\begin{verbatim}
Input: x = 2.10000, n = 3
Output: 9.26100
Explanation: 2.1^3 = 9.261
\end{verbatim}

\textbf{Example 3:}

\begin{verbatim}
Input: x = 2.00000, n = -2
Output: 0.25000
Explanation: 2^-2 = 1/4 = 0.25
\end{verbatim}

\textbf{Constraints:}

\begin{itemize}
    \item \(-100.0 < x < 100.0\)
    \item \(-2^{31} \leq n \leq 2^{31} - 1\)
    \item \(-10^4 \leq x^n \leq 10^4\)
\end{itemize}


\section*{Algorithmic Approach}

To efficiently compute \(x^n\), especially for large values of \(n\), we can utilize the **Exponentiation by Squaring** technique. This method reduces the time complexity from linear \(O(n)\) to logarithmic \(O(\log n)\), making it highly efficient for large exponents.

\subsection*{1. Handling Negative Exponents}

If \(n\) is negative, we can compute \(x^{-n} = \frac{1}{x^n}\). This transformation allows us to handle negative exponents seamlessly using the same exponentiation logic.

\subsection*{2. Exponentiation by Squaring}

The core idea is to break down the exponentiation process into smaller, manageable parts:
\begin{enumerate}
    \item **Base Case:** If \(n = 0\), return 1 as any number raised to the power of 0 is 1.
    \item **Recursive Case:**
    \begin{itemize}
        \item If \(n\) is even, compute \((x^{2})^{n/2}\).
        \item If \(n\) is odd, compute \(x \times x^{n-1}\).
    \end{itemize}
\end{enumerate}

This approach minimizes the number of multiplications required, leading to significant performance gains.

\marginnote{Exponentiation by Squaring reduces the number of multiplications, achieving \(O(\log n)\) time complexity.}

\section*{Complexities}

\begin{itemize}
    \item \textbf{Time Complexity:} \(O(\log |n|)\). The algorithm reduces the exponent by half in each step, leading to logarithmic time complexity.
    
    \item \textbf{Space Complexity:} \(O(1)\). The iterative approach uses constant space, whereas the recursive approach may use \(O(\log |n|)\) space due to recursion stack.
\end{itemize}

\section*{Python Implementation}

\marginnote{Implementing exponentiation by squaring ensures an optimal solution with logarithmic time complexity and constant space usage.}

Below is the complete Python code implementing the \texttt{myPow} function using the Exponentiation by Squaring approach:

\begin{fullwidth}
\begin{lstlisting}[language=Python]
class Solution:
    def myPow(self, x: float, n: int) -> float:
        if n == 0:
            return 1.0
        if n < 0:
            x = 1 / x
            n = -n
        result = 1.0
        current_product = x
        while n > 0:
            if n % 2 == 1:
                result *= current_product
            current_product *= current_product
            n //= 2
        return result

# Example usage:
solution = Solution()
print(solution.myPow(2.0, 10))     # Output: 1024.0
print(solution.myPow(2.1, 3))      # Output: 9.261
print(solution.myPow(2.0, -2))     # Output: 0.25
\end{lstlisting}
\end{fullwidth}

This implementation handles both positive and negative exponents efficiently. It iteratively squares the base and multiplies it to the result when the current exponent bit is set.

\section*{Explanation}

The \texttt{myPow} function efficiently computes \(x^n\) using the Exponentiation by Squaring method. Here's a detailed breakdown of the implementation:

\subsection*{Handling Negative Exponents}

\begin{enumerate}
    \item **Negative Exponent Transformation:**
    \begin{itemize}
        \item If \(n\) is negative, we transform the problem to compute \(\frac{1}{x^n}\) by setting \(x = \frac{1}{x}\) and \(n = -n\).
    \end{itemize}
\end{enumerate}

\subsection*{Exponentiation by Squaring}

\begin{enumerate}
    \item **Initialization:**
    \begin{itemize}
        \item \texttt{result} is initialized to 1.0, which will accumulate the final power value.
        \item \texttt{current\_product} is initialized to \(x\), representing the current power of \(x\) being considered.
    \end{itemize}
    
    \item **Iterative Processing:**
    \begin{itemize}
        \item While \(n > 0\):
        \begin{itemize}
            \item **Odd Exponent:** If \(n\) is odd (\(n \% 2 == 1\)), multiply \texttt{result} by \texttt{current\_product}.
            \item **Square the Base:** Regardless of whether \(n\) is odd or even, square \texttt{current\_product} to prepare for the next iteration.
            \item **Halve the Exponent:** Divide \(n\) by 2 using integer division to move to the next bit.
        \end{itemize}
    \end{itemize}
    
    \item **Final Result:**
    \begin{itemize}
        \item After processing all bits of \(n\), \texttt{result} contains the value of \(x^n\), which is returned.
    \end{itemize}
\end{enumerate}

\subsection*{Example Walkthrough}

Consider \(x = 2.0\) and \(n = 10\):

\begin{itemize}
    \item **Initial State:**
    \begin{itemize}
        \item \texttt{result} = 1.0
        \item \texttt{current\_product} = 2.0
        \item \(n = 10\)
    \end{itemize}
    
    \item **First Iteration (\(n = 10\)):**
    \begin{itemize}
        \item \(10\) is even.
        \item \texttt{current\_product} = \(2.0^2 = 4.0\)
        \item \(n = 5\)
    \end{itemize}
    
    \item **Second Iteration (\(n = 5\)):**
    \begin{itemize}
        \item \(5\) is odd.
        \item \texttt{result} = \(1.0 \times 4.0 = 4.0\)
        \item \texttt{current\_product} = \(4.0^2 = 16.0\)
        \item \(n = 2\)
    \end{itemize}
    
    \item **Third Iteration (\(n = 2\)):**
    \begin{itemize}
        \item \(2\) is even.
        \item \texttt{current\_product} = \(16.0^2 = 256.0\)
        \item \(n = 1\)
    \end{itemize}
    
    \item **Fourth Iteration (\(n = 1\)):**
    \begin{itemize}
        \item \(1\) is odd.
        \item \texttt{result} = \(4.0 \times 256.0 = 1024.0\)
        \item \texttt{current\_product} = \(256.0^2 = 65536.0\)
        \item \(n = 0\)
    \end{itemize}
    
    \item **Termination:**
    \begin{itemize}
        \item \(n = 0\), exit loop.
        \item \texttt{result} = 1024.0
    \end{itemize}
\end{itemize}

Thus, \(2.0^{10} = 1024.0\) is correctly computed.

\section*{Why This Approach}

The **Exponentiation by Squaring** method is chosen for its optimal efficiency and simplicity. Compared to the naive approach of multiplying \(x\) by itself \(n\) times (\(O(n)\) time complexity), this method reduces the number of multiplications to \(O(\log n)\), making it highly efficient for large exponents. Additionally, by handling negative exponents through reciprocal transformation, the algorithm seamlessly accommodates all integer values of \(n\).

\section*{Alternative Approaches}

While Exponentiation by Squaring is the most efficient method for this problem, there are alternative ways to compute \(x^n\):

\subsection*{1. Naive Iterative Approach}

Multiply \(x\) by itself \(n\) times.

\begin{lstlisting}[language=Python]
class Solution:
    def myPow(self, x: float, n: int) -> float:
        if n == 0:
            return 1.0
        if n < 0:
            x = 1 / x
            n = -n
        result = 1.0
        for _ in range(n):
            result *= x
        return result
\end{lstlisting}

\textbf{Complexities:}
\begin{itemize}
    \item \textbf{Time Complexity:} \(O(n)\)
    \item \textbf{Space Complexity:} \(O(1)\)
\end{itemize}

\subsection*{2. Recursive Exponentiation by Squaring}

Implement the exponentiation by squaring technique using recursion.

\begin{lstlisting}[language=Python]
class Solution:
    def myPow(self, x: float, n: int) -> float:
        if n == 0:
            return 1.0
        if n < 0:
            x = 1 / x
            n = -n
        half = self.myPow(x, n // 2)
        if n % 2 == 0:
            return half * half
        else:
            return half * half * x
\end{lstlisting}

\textbf{Complexities:}
\begin{itemize}
    \item \textbf{Time Complexity:} \(O(\log n)\)
    \item \textbf{Space Complexity:} \(O(\log n)\) due to recursion stack
\end{itemize}

\subsection*{3. Built-In Functions}

Use language-specific built-in functions for exponentiation.

\begin{lstlisting}[language=Python]
class Solution:
    def myPow(self, x: float, n: int) -> float:
        return x ** n
\end{lstlisting}

\textbf{Note:} While this method is straightforward, it may not provide insights into the underlying algorithmic process and may have different performance characteristics based on the language implementation.

\section*{Similar Problems to This One}

Several problems involve exponentiation and power-related computations, utilizing similar algorithmic strategies:

\begin{itemize}
    \item \textbf{Sqrt(x)}: Compute the square root of a number using binary search or Newton's method.
    \item \textbf{Integer to English Words}: Convert an integer to its English words representation.
    \item \textbf{Happy Number}: Determine if a number is a "happy" number based on specific iterative processes.
    \item \textbf{Power of Three}: Check if a number is a power of three using logarithms or iterative multiplication.
    \item \textbf{Nth Fibonacci Number}: Compute the \(n^{th}\) Fibonacci number using dynamic programming or matrix exponentiation.
    \item \textbf{Modular Exponentiation}: Compute \(x^n \mod m\) efficiently using exponentiation by squaring.
\end{itemize}

These problems reinforce the concepts of efficient exponentiation, binary search, dynamic programming, and iterative computation in various contexts.

\section*{Things to Keep in Mind and Tricks}

When working with exponentiation and power-related algorithms, consider the following tips and best practices to enhance efficiency and correctness:

\begin{itemize}
    \item \textbf{Exponentiation by Squaring}: Utilize this technique to reduce the number of multiplications from \(O(n)\) to \(O(\log n)\).
    \index{Exponentiation by Squaring}
    
    \item \textbf{Handling Negative Exponents}: Transform negative exponents into their reciprocal form to simplify computations.
    \index{Negative Exponents}
    
    \item \textbf{Avoid Floating-Point Precision Issues}: Be cautious with floating-point arithmetic, especially when dealing with large exponents or very small/large results.
    \index{Floating-Point Precision}
    
    \item \textbf{Recursive vs. Iterative Implementations}: Choose between recursive and iterative approaches based on space constraints and language-specific recursion limits.
    \index{Recursive vs. Iterative}
    
    \item \textbf{Edge Cases}: Always consider edge cases such as \(n = 0\), \(n = 1\), \(x = 0\), and negative values of \(x\) and \(n\).
    \index{Edge Cases}
    
    \item \textbf{Optimizing Space}: Prefer iterative solutions to avoid additional space usage associated with recursion.
    \index{Space Optimization}
    
    \item \textbf{Understanding Language-Specific Operators}: Familiarize yourself with how your programming language handles exponentiation and bitwise operations.
    \index{Language-Specific Operators}
    
    \item \textbf{Use of Helper Functions}: Create helper functions for repetitive tasks, such as handling negative exponents or squaring the base.
    \index{Helper Functions}
    
    \item \textbf{Mathematical Properties}: Leverage mathematical identities and properties to simplify computations, such as \(x^{a+b} = x^a \times x^b\).
    \index{Mathematical Properties}
    
    \item \textbf{Testing Thoroughly}: Implement comprehensive test cases covering all possible scenarios, including large exponents and edge conditions.
    \index{Testing}
\end{itemize}

\section*{Corner and Special Cases to Test When Writing the Code}

When implementing the \texttt{myPow} function, it's crucial to consider and rigorously test various edge cases to ensure robustness and correctness:

\begin{itemize}
    \item \textbf{Zero Exponent (\(n = 0\))}: Any number raised to the power of zero should return 1.
    \index{Zero Exponent}
    
    \item \textbf{One Exponent (\(n = 1\))}: The function should return the base itself.
    \index{One Exponent}
    
    \item \textbf{Negative Exponents (\(n < 0\))}: Ensure that the function correctly computes the reciprocal of the power.
    \index{Negative Exponents}
    
    \item \textbf{Base Zero (\(x = 0\))}: Handle cases where the base is zero, considering different exponents.
    \index{Base Zero}
    
    \item \textbf{Large Exponents}: Test with very large values of \(n\) to verify that the algorithm maintains efficiency and accuracy.
    \index{Large Exponents}
    
    \item \textbf{Small Exponents}: Test with small exponents, including \(n = -1\), to ensure correct handling.
    \index{Small Exponents}
    
    \item \textbf{Base One (\(x = 1\)) and Base Negative One (\(x = -1\))}: These cases have predictable outcomes regardless of the exponent's magnitude.
    \index{Base One and Negative One}
    
    \item \textbf{Fractional Bases}: Ensure that the function accurately computes powers for non-integer bases.
    \index{Fractional Bases}
    
    \item \textbf{Maximum and Minimum Integer Values}: Test with \(n = 2^{31} - 1\) and \(n = -2^{31}\) to ensure no overflow or underflow occurs.
    \index{Maximum and Minimum Integer Values}
    
    \item \textbf{Non-Integer Exponents}: Although the problem specifies integer exponents, ensure that the function handles unexpected input gracefully if possible.
    \index{Non-Integer Exponents}
\end{itemize}

\section*{Implementation Considerations}

When implementing the \texttt{myPow} function, keep in mind the following considerations to ensure robustness and efficiency:

\begin{itemize}
    \item \textbf{Data Type Selection}: Use appropriate data types that can handle large exponents and avoid integer overflow or floating-point precision issues.
    \index{Data Type Selection}
    
    \item \textbf{Optimizing Bitwise Operations}: Ensure that bitwise operations are used efficiently without introducing unnecessary computations or steps.
    \index{Optimizing Bitwise Operations}
    
    \item \textbf{Language-Specific Limitations}: Be aware of how your programming language handles large integers, floating-point arithmetic, and recursion limits.
    \index{Language-Specific Limitations}
    
    \item \textbf{Avoiding Redundant Calculations}: In the exponentiation by squaring method, ensure that each multiplication is necessary and contributes to the final result.
    \index{Avoiding Redundant Calculations}
    
    \item \textbf{Code Readability and Documentation}: Maintain clear and readable code through meaningful variable names and comprehensive comments, especially when dealing with complex bitwise operations.
    \index{Code Readability}
    
    \item \textbf{Edge Case Handling}: Implement checks and balances to handle edge cases appropriately, preventing incorrect results or runtime errors.
    \index{Edge Case Handling}
    
    \item \textbf{Testing and Validation}: Develop a comprehensive suite of test cases that cover all possible scenarios, including edge cases, to validate the correctness and efficiency of the implementation.
    \index{Testing and Validation}
    
    \item \textbf{Scalability}: Design the algorithm to scale efficiently with increasing input sizes, maintaining performance and resource utilization.
    \index{Scalability}
    
    \item \textbf{Floating-Point Precision}: Be cautious with floating-point arithmetic, especially when dealing with very large or very small numbers, to avoid precision loss.
    \index{Floating-Point Precision}
    
    \item \textbf{Using Built-In Functions}: Where appropriate, leverage built-in functions or libraries that can perform exponentiation more efficiently.
    \index{Built-In Functions}
    
    \item \textbf{Iterative vs. Recursive Implementations}: Prefer iterative solutions for problems where recursion might lead to stack overflow or increased space complexity.
    \index{Iterative vs. Recursive}
\end{itemize}

\section*{Conclusion}

The \textbf{Pow(x, n)} problem exemplifies the application of mathematical algorithms and bit manipulation techniques to solve computational challenges efficiently. By leveraging the Exponentiation by Squaring method, the algorithm achieves logarithmic time complexity and constant space usage, making it highly suitable for large exponents. Understanding and implementing such techniques not only enhances problem-solving skills but also provides a foundation for tackling a wide array of mathematical and algorithmic problems in computer science. Mastery of these concepts is essential for optimizing performance in software development, scientific computing, and various other fields that rely on efficient numerical computations.

\printindex

% \input{sections/mathematical_algorithms}
% \input{sections/pow_x_n}
% \input{sections/sqrt_x}
% \input{sections/integer_to_english_words}
% \input{sections/happy_number}